---
title: "Project"
author: "Dhaen Alkaabi    Aidan Pierre-Louis    Luke Marini   Noah Eisenberg"
output: html_document
date: "2023-12-11"
---

#### https://github.com/nflverse/nflverse-data

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(openxlsx)
library(tidyverse)
library(kableExtra)
library(dplyr)
library(tidyr)
library(grid)
library(gridExtra)
library(modelr)
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)

#draft_data <- read.xlsx("C:\\STAT240\\data\\draft_picks.xlsx") deprecating due to absolutepath
#recommend to set working directory to current file location

draft_data <- read.xlsx("../data/draft_picks.xlsx")
player_data <- read_csv("../data/combine.csv")

player_data = player_data %>% rename(pfr_player_id= 'pfr_id', cfb_player_id= 'cfb_id', pfr_player_name= 'player_name')
merge_data = right_join(player_data, draft_data)
```

# Research Question 1.1: Do draft pick number and various player draft performance attributes, such as rushing yards, recieving yards, and defensive sacks exhibit a significant correlation towards a player's success in the NFL and draft? And if so, which positions have the strongest and weakest correlations?

```{r}
#Data Cleaning and Wrangling

draft_data_clean <- draft_data %>%
  select(season, pick, position, rush_yards, rec_yards, def_sacks) %>%
  mutate(season = as.numeric(season)) %>% 
  filter(!is.na(pick)) %>%
  na.omit() 
```

> The first segment of the code is dedicated to preparing the dataset for analysis. It focuses on selecting the relevant columns, converting the season to a numeric format, filtering out rows with missing pick values, and removing rows with any NA values. This step is crucial as it lays the foundation for accurate and meaningful analysis. By ensuring the data is clean and relevant, this process directly supports the investigation of the research question, particularly in examining the relationship between draft pick numbers and player success metrics. Clean and well-structured data is vital for deriving valid conclusions about how draft picks relate to player performance in the NFL.



```{r}
#Data Visualization

ggplot(draft_data_clean, aes(x = pick, y = rush_yards)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~position) +
  theme_minimal() +
  labs(title = "Draft Pick vs Rush Yards by Position")
```



> The plot is a collection of scatterplots, each representing a different position in the NFL, comparing draft pick numbers (x-axis) to rushing yards (y-axis). The points represent individual players, and the blue lines show the trend of rushing yards against draft picks within each position. Positions like DB and WR show a slight negative trend, suggesting higher draft picks typically have more rushing yards. Most positions, especially non-rushing ones like T and DT, show little to no trend, indicating draft pick number doesn't strongly correlate with rushing yards for those positions.

> This code creates a scatter plot with a linear regression line to visually explore the relationship between the draft pick number (pick) and rushing yards (rush_yards), one of the success metrics. It also uses facet_wrap to separate the plots by player position, allowing for an easy comparison across different roles. This visualization is integral to understanding the correlation between draft pick number and player success, specifically how this relationship might vary across different positions. It provides a clear visual representation to address the first part of the research question, focusing on the significance of the correlation between draft pick number, player success metrics, and variations across different positions.

```{r}
#Statistical Modeling: Correlation and Regression Analysis

cor_pick_rush <- cor.test(draft_data_clean$pick, draft_data_clean$rush_yards, use = "complete.obs")
cor_pick_rec <- cor.test(draft_data_clean$pick, draft_data_clean$rec_yards, use = "complete.obs")
cor_pick_def_sacks <- cor.test(draft_data_clean$pick, draft_data_clean$def_sacks, use = "complete.obs")

# Print the results of the correlation tests
print(paste("Correlation between draft pick and rushing yards: ", cor_pick_rush$estimate))
print(paste("P-value:", cor_pick_rush$p.value))
print(paste("Correlation between draft pick and receiving yards: ", cor_pick_rec$estimate))
print(paste("P-value:", cor_pick_rec$p.value))
print(paste("Correlation between draft pick and defensive sacks: ", cor_pick_def_sacks$estimate))
print(paste("P-value:", cor_pick_def_sacks$p.value))

# Fit a linear regression model for rush yards vs draft pick
model_rush_yards <- lm(rush_yards ~ pick, data = draft_data_clean)

# Print the summary of the regression model
summary(model_rush_yards)
```

> The correlation tests show a very weak negative correlation between draft pick number and both rushing and receiving yards, which is not statistically significant. However, there is a moderate and statistically significant negative correlation between draft pick number and defensive sacks. The linear regression model for predicting rushing yards from the draft pick number is not statistically significant and explains none of the variability in rushing yards.

> The analysis directly addresses the research question by examining if the draft pick number correlates with measurable success in the NFL, such as rushing yards, receiving yards, and defensive sacks. While a significant negative correlation with defensive sacks suggests that lower draft picks (which are considered more premier) tend to perform better in this metric, the lack of significant correlations with rushing and receiving yards implies that the draft pick number may not be a consistent predictor of success for these metrics across various positions. This indicates that other player attributes might also need to be considered to fully assess a player's potential success in the NFL.

# Research Question 1.2: Do various player attributes, such as height, weight, and speed, exhibit a significant correlation towards a player's success in the draft pick order? And if so, which positions have the strongest and weakest correlations?


```{r}
merge_attrs = merge_data %>% filter(!is.na(pick)) %>%
  select(season, round, pick, team, pfr_player_name, position, age, ht, wt, forty, bench, vertical, broad_jump, cone, shuttle)
```
```{r, eval = FALSE}
attr_plot <- function(attr) {
merge_attrs %>% ggplot(aes(x= pick,y= attr)) +
  geom_point() +
  geom_smooth()
}

attrs <- list(age, ht, wt, forty, bench, vertical, broad_jump, cone, shuttle)
attr_plots = lapply(attrs, attr_plot)
```

``` {r}
merge_attrs %>% filter(!is.na(age)) %>% ggplot(aes(x= pick,y= age)) +
  geom_point(alpha=0.5) +
  geom_smooth(se=FALSE,method="lm",color="red") + geom_smooth(se=FALSE,method="loess",color="blue")

age_lm = lm(pick ~ age,data=merge_attrs %>% filter(!is.na(age)))
merge_attrs %>% filter(!is.na(age)) %>%
  add_residuals(age_lm) %>%
  add_predictions(age_lm) %>%
  ggplot(aes(x= age,y=resid)) + 
  geom_point(alpha=0.5) + 
  geom_hline(yintercept = 0, color="red") + 
  geom_smooth(se=FALSE,method="loess",color="blue")


merge_attrs %>% filter(!is.na(wt)) %>% ggplot(aes(x= pick,y= wt)) +
  geom_point(alpha=0.5) +
  geom_smooth(se=FALSE,method="lm",color="red") + geom_smooth(se=FALSE,method="loess",color="blue")

wt_lm = lm(pick ~ wt,data=merge_attrs %>% filter(!is.na(wt)))
merge_attrs %>% filter(!is.na(wt)) %>%
  add_residuals(wt_lm) %>%
  add_predictions(wt_lm) %>%
  ggplot(aes(x= wt,y=resid)) + 
  geom_point(alpha=0.5) + 
  geom_hline(yintercept = 0, color="red") + 
  geom_smooth(se=FALSE,method="loess",color="blue")


merge_attrs %>% filter(!is.na(forty)) %>% ggplot(aes(x= pick,y= forty)) +
  geom_point(alpha=0.5) +
  geom_smooth(se=FALSE,method="lm",color="red") + geom_smooth(se=FALSE,method="loess",color="blue")

ht_lm = lm(pick ~ ht,data=merge_attrs %>% filter(!is.na(ht)))
merge_attrs %>% filter(!is.na(ht)) %>%
  add_residuals(ht_lm) %>%
  add_predictions(ht_lm) %>%
  ggplot(aes(x= ht,y=resid)) + 
  geom_point(alpha=0.5) + 
  geom_hline(yintercept = 0, color="red") + 
  geom_smooth(se=FALSE,method="loess",color="blue")


merge_attrs %>% filter(!is.na(bench)) %>% ggplot(aes(x= pick,y= bench)) +
  geom_point(alpha=0.5) +
  geom_smooth(se=FALSE,method="lm",color="red") + geom_smooth(se=FALSE,method="loess",color="blue")

bench_lm = lm(pick ~ bench,data=merge_attrs %>% filter(!is.na(bench)))
merge_attrs %>% filter(!is.na(bench)) %>%
  add_residuals(bench_lm) %>%
  add_predictions(bench_lm) %>%
  ggplot(aes(x= bench,y=resid)) + 
  geom_point(alpha=0.5) + 
  geom_hline(yintercept = 0, color="red") + 
  geom_smooth(se=FALSE,method="loess",color="blue")


merge_attrs %>% filter(!is.na(vertical)) %>% ggplot(aes(x= pick,y= vertical)) +
  geom_point(alpha=0.5) +
  geom_smooth(se=FALSE,method="lm",color="red") + geom_smooth(se=FALSE,method="loess",color="blue")

v_lm = lm(pick ~ vertical,data=merge_attrs %>% filter(!is.na(vertical)))
merge_attrs %>% filter(!is.na(vertical)) %>%
  add_residuals(v_lm) %>%
  add_predictions(v_lm) %>%
  ggplot(aes(x= vertical,y=resid)) + 
  geom_point(alpha=0.5) + 
  geom_hline(yintercept = 0, color="red") + 
  geom_smooth(se=FALSE,method="loess",color="blue")


merge_attrs %>% filter(!is.na(broad_jump)) %>% ggplot(aes(x= pick,y= broad_jump)) +
  geom_point(alpha=0.5) +
  geom_smooth(se=FALSE,method="lm",color="red") + geom_smooth(se=FALSE,method="loess",color="blue")

bj_lm = lm(pick ~ broad_jump,data=merge_attrs %>% filter(!is.na(broad_jump)))
merge_attrs %>% filter(!is.na(broad_jump)) %>%
  add_residuals(bj_lm) %>%
  add_predictions(bj_lm) %>%
  ggplot(aes(x= broad_jump,y=resid)) + 
  geom_point(alpha=0.5) + 
  geom_hline(yintercept = 0, color="red") + 
  geom_smooth(se=FALSE,method="loess",color="blue")


merge_attrs %>% filter(!is.na(cone)) %>% ggplot(aes(x= pick,y= cone)) +
  geom_point(alpha=0.5) +
  geom_smooth(se=FALSE,method="lm",color="red") + geom_smooth(se=FALSE,method="loess",color="blue")

c_lm = lm(pick ~ cone,data=merge_attrs %>% filter(!is.na(cone)))
merge_attrs %>% filter(!is.na(cone)) %>%
  add_residuals(c_lm) %>%
  add_predictions(c_lm) %>%
  ggplot(aes(x= cone,y=resid)) + 
  geom_point(alpha=0.5) + 
  geom_hline(yintercept = 0, color="red") + 
  geom_smooth(se=FALSE,method="loess",color="blue")


merge_attrs %>% filter(!is.na(shuttle)) %>% ggplot(aes(x= pick,y= shuttle)) +
  geom_point(alpha=0.5) +
  geom_smooth(se=FALSE,method="lm",color="red") + geom_smooth(se=FALSE,method="loess",color="blue")

s_lm = lm(pick ~ shuttle,data=merge_attrs %>% filter(!is.na(shuttle)))
merge_attrs %>% filter(!is.na(shuttle)) %>%
  add_residuals(s_lm) %>%
  add_predictions(s_lm) %>%
  ggplot(aes(x= shuttle,y=resid)) + 
  geom_point(alpha=0.5) + 
  geom_hline(yintercept = 0, color="red") + 
  geom_smooth(se=FALSE,method="loess",color="blue")
```

>TODO Conclusion drawn from plots - Aidan

# Research Question 2. Are there any specific draft positions that are more likely to produce successful players than others?

```{r}
#Inference on Proportions

success_threshold <- median(draft_data_clean$rush_yards, na.rm = TRUE)
draft_data_clean$successful_player <- ifelse(draft_data_clean$rush_yards > success_threshold, 1, 0)


success_proportion <- draft_data_clean %>%
  group_by(position) %>%
  summarise(
    proportion_success = mean(successful_player),
    n = n()
  )

print(success_proportion)
```


> The table lists NFL positions along with the proportion of players at each position who exceeded the median rushing yards, used as a criterion for success. For positions with a higher number of players (like DB and LB), there's a greater chance of players exceeding the success threshold. Positions with a proportion of zero indicate no players exceeded the median rushing yards, which may reflect the typical role of the position in the game. Positions with only one player (like G, T, TE, WR) provide limited data and thus, less reliable conclusions about success rates.

> This part of the code defines a success criterion based on rushing yards and calculates the proportion of successful players in each position. It is instrumental in examining whether certain positions are more likely to yield successful players, thus addressing the latter part of your research question. By calculating these proportions, this analysis provides a statistical foundation to infer position-specific success rates, enhancing the understanding of how different positions correlate with the success rates of players in the NFL.

> Here I will dive in to the correlation between draft pick and success for the four main offensive skill positions in the NFL: Quarterback (QB), Running Back (RB), Wide Reciever (WR), and Tight End (TE). To create a metric to compute and compare success, I am utilizing fantasy points, a scoring system that is commonly used to calcuate "points" based on a player's statistics. Below I evaluate if draft pick can be used to explain fantasy points using a linear regression model, while using a logarithmic transformation on fantasy points.

```{r}
draft_data = draft_data %>% replace(is.na(.), 0)
qb = draft_data %>%
  filter(position == "QB") %>%
  select(round, pick, position, pass_yards, pass_tds, pass_ints, rush_yards, rush_tds, probowls, allpro) %>%
  mutate(fantasy_points = pass_yards*0.25 + pass_tds*4 + pass_ints*-2 + rush_yards*0.1 + rush_tds*6) %>%
  mutate(
         log_fpoints = case_when(
           fantasy_points <= 0 ~ 0, 
           fantasy_points != 0 ~ log(fantasy_points)))

rb = draft_data %>%
  filter(position == "RB") %>%
  select(round, pick, rush_yards, rush_tds, probowls, allpro, seasons_started) %>%
  mutate(fantasy_points = rush_yards*0.1 + rush_tds*6) %>%
  mutate(
         log_fpoints = case_when(
           fantasy_points <= 0 ~ 0, 
           fantasy_points != 0 ~ log(fantasy_points)))

wr = draft_data %>%
  filter(position == "WR") %>%
  select(round, pick, rec_yards, rec_tds, receptions, probowls, allpro, seasons_started) %>%
  mutate(fantasy_points = receptions*0.5 + rec_yards*0.1 + rec_tds*6) %>%
  mutate(
         log_fpoints = case_when(
           fantasy_points <= 0 ~ 0, 
           fantasy_points != 0 ~ log(fantasy_points)))


te = draft_data %>%
  filter(position == "TE") %>%
  select(round, pick, rec_yards, rec_tds, receptions, probowls, allpro, seasons_started) %>%
  mutate(fantasy_points = receptions*0.5 + rec_yards*0.1 + rec_tds*6) %>%
  mutate(
         log_fpoints = case_when(
           fantasy_points <= 0 ~ 0, 
           fantasy_points != 0 ~ log(fantasy_points)))

qb_cor = cor(qb$pick, qb$log_fpoints)
qb_lm = lm(log_fpoints ~ pick, data = qb)


qb_plot = ggplot(qb, aes(x = pick, y = log_fpoints)) + geom_point() + geom_smooth(se = FALSE, method = "lm") + xlab("Draft Pick") + ylab("Fantasy Points (log scale)") + ggtitle("Quarterbacks (r = -0.64)")

rb_cor = cor(rb$pick, rb$log_fpoints)
rb_lm = lm(log_fpoints ~ pick, data = rb)


rb_plot = ggplot(rb, aes(x = pick, y = log_fpoints)) + geom_point() + geom_smooth(se = FALSE, method = "lm") + xlab("Draft Pick") + ylab("Fantasy Points (log scale)") + ggtitle("Running Backs (r = -0.61)")


wr_cor = cor(wr$pick, wr$log_fpoints)
wr_lm = lm(log_fpoints ~ pick, data = wr)


wr_plot = ggplot(wr, aes(x = pick, y = log_fpoints)) + geom_point() + geom_smooth(se = FALSE, method = "lm") + xlab("Draft Pick") + ylab("Fantasy Points (log scale)") + ggtitle("Wide Recievers (r = -0.60)")

te_cor = cor(te$pick, te$log_fpoints)
te_lm = lm(log_fpoints ~ pick, data = te)


te_plot = ggplot(te, aes(x = pick, y = log_fpoints)) + geom_point() + geom_smooth(se = FALSE, method = "lm") + xlab("Draft Pick") + ylab("Fantasy Points (log scale)") + ggtitle("Tight Ends (r = -0.56)")

tibble(qb_cor, rb_cor, wr_cor, te_cor)

grid.arrange(qb_plot, rb_plot, wr_plot, te_plot, top = textGrob("NFL Players: Draft Pick vs Fantasy Points by Position"))
```

> The graphs show a similar trend among all four positions: A moderate negative correlation between draft pick and log(fantasy points), indicating that a higher draft pick does seem to correlate with more success in the NFL.

```{r}
qb = qb %>%
  add_residuals(qb_lm)

qb_resid = ggplot(qb, aes(x = pick, y = resid)) + geom_point() + geom_hline(yintercept = 0, color = "red", linetype = "dashed") + xlab("Draft Pick") + ylab("Residuals") + ggtitle("Quarterbacks") + geom_smooth(se=FALSE)

rb = rb %>%
  add_residuals(rb_lm)

rb_resid = ggplot(rb, aes(x = pick, y = resid)) + geom_point() + geom_hline(yintercept = 0, color = "red", linetype = "dashed") + xlab("Draft Pick") + ylab("Residuals") + ggtitle("Running Backs")+ geom_smooth(se=FALSE)

wr = wr %>%
  add_residuals(wr_lm)

wr_resid = ggplot(wr, aes(x = pick, y = resid)) + geom_point() + geom_hline(yintercept = 0, color = "red", linetype = "dashed") + xlab("Draft Pick") + ylab("Residuals") + ggtitle("Wide Receivers") + geom_smooth(se=FALSE)

te = te %>%
  add_residuals(qb_lm)

te_resid = ggplot(te, aes(x = pick, y = resid)) + geom_point() + geom_hline(yintercept = 0, color = "red", linetype = "dashed") + xlab("Draft Pick") + ylab("Residuals") + ggtitle("Tight Ends") + geom_smooth(se=FALSE)

grid.arrange(qb_resid, rb_resid, wr_resid, te_resid, top = textGrob("Residual Plots"))
```

 > The residual plots showcase a slight but consistent pattern that include positive residuals near the boundaries and negative residuals in near the middle, but overall the pattern does not seem too signficant and the linear model seems reasonable.


> 

> Research Question 3. Can we develop a predictive model to forecast a player's NFL success based on their draft position and other relevant factors?

```{r}
small_value <- 1e-6


draft_data_clean <- draft_data_clean %>%
  mutate(
    rush_yards = ifelse(rush_yards <= 0, log1p(small_value), log1p(rush_yards)),
    predicted_rush_yards = ifelse(rush_yards <= 0, log1p(small_value), log1p(rush_yards))
  )


draft_data_clean <- na.omit(draft_data_clean)


plot_data <- draft_data_clean %>%
  select(pick, rush_yards, predicted_rush_yards) %>%
  pivot_longer(cols = c("rush_yards", "predicted_rush_yards"),
               names_to = "type",
               values_to = "yards")


ggplot(plot_data, aes(x = pick, y = yards, color = type)) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Observed vs Predicted Rushing Yards by Draft Pick",
       x = "Draft Pick",
       y = "Rushing Yards (log scale)")
```

---
title: "SignificantResults"
output: html_document
date: "2023-12-05"
---

```{r setup2, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# THIS CODE CHUNK CREATES THE DATA SET WE ARE GOING TO WORK WITH TO LOOK AT CORRELATIONS

combine = read_csv("../data/combine.csv")
players = read_csv("../data/draft_picks.csv")

combine = combine %>%
  select(draft_year, draft_team, draft_round, draft_ovr, player_name, pos, ht, wt, forty) %>%
  drop_na() 

players = players %>%
  select(season, round, pick, team, pfr_player_name, position, age, hof, allpro, probowls) %>%
  drop_na() %>%
  filter(season >= 2000) 

names(players) = c("season", "round", "pick", "team", "player", "position", "age", "HOF", "AP", "PB")
names(combine) = c("season", "team", "round", "pick", "player", "position", "height", "weight", "speed")

players = players %>%
  select(player, season, round, pick, position, age, HOF, AP, PB) %>%
  drop_na()

combine = combine %>%
  select(player, season, round, pick, position, height, weight, speed) %>%
  drop_na()

data = players %>%
  left_join(combine, by = c("player", "season", "round", "pick", "position")) %>%
  drop_na() %>%
  distinct() %>%
  filter(position == 'WR' | position == 'QB' | position == 'TE' | position == 'RB') 

data = data %>%
  mutate(HOFscore = case_when(
    HOF == 'FALSE' ~ 0,
    HOF == 'TRUE' ~ 10
  )) %>%
  mutate(success_score = HOFscore + 3 * AP + PB) %>%
  select(everything(), -HOFscore)
```


> SIGNIFICANT RESULTS FOR QUARTERBACKS:

```{r}
# looking at proportions of successful qbs based on whether they were drafted top100 or not, and top32 or not
quarterbacks = data %>%
  filter(position == 'QB') 

top100 = quarterbacks %>%
  filter(pick <= 100) %>%
  mutate(successful = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))

notTop100 = quarterbacks %>%
  filter(pick > 100) %>%
  mutate(successful = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))

ggplot(top100, aes(x = successful)) + geom_bar() + ggtitle("Successful vs. unsuccessful quarterbacks drafted top 100")
ggplot(notTop100, aes(x = successful)) + geom_bar() + ggtitle("Successful vs. unsuccessful quarterbacks drafted outside the top 100")

firstRound = quarterbacks %>%
  filter(pick <= 32) %>%
  mutate(successful = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))

notFirstRound = quarterbacks %>%
  filter(pick > 32) %>%
  mutate(successful = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))

ggplot(firstRound, aes(x = successful)) + geom_bar() + ggtitle("Successful vs. unsuccessful quarterbacks drafted in the first round")
ggplot(notFirstRound, aes(x = successful)) + geom_bar() + ggtitle("Successful vs. unsuccessful quarterbacks drafted outside the first round")
```

```{r}
#testing for significance for only first round qbs: Ho: Psuccess = 0.096, Ha: Psuccess > 0.096
#this can be considered binomial because there are only 2 options: successful or not, each qb's success is not determined by any other qb's success, sample size is pre-specified, and we are assuming each first round qb has the same probability of becoming successful

n = 57
p = 0.096
mean = n * p
sd = sqrt(mean * (1-p))

pval = (1 - pnorm(13, mean, sd))
pval

n = 107
p = 0.096
mean = n * p
sd = sqrt(mean * (1-p))

pval = (1 - pnorm(17, mean, sd))
pval
```

> A large portion of successful quarterbacks have been drafted earlier. Out of 219 quarterbacks, 21 have been considered successul (9.6%). 

>48.9% of QBs were selected in the top 100, and 16.8% of them were successful.
>26% of QBs were selected in the first round, and 24.6% of them were successful.

>There is extremely strong evidence (p value = 0.0004, one-sided binomial test) that in the long run a quarterback drafted in the first round is more likely to be successful than a quarterback not drafted in the first round.

>There is also strong evidence (p value = 0.014, one-sided binomial test) that in the long run a quarterback drafted in the top 100 picks is more likely to be successful than a quarterback not drafted in the top 100.


> SIGNIFICANT RESULTS FOR WIDE RECEIVERS:

```{r}
wr = data %>%
  filter(position == 'WR')

firstRoundWR = wr %>%
  filter(pick <= 32)

#ggplot(wr, aes(x = weight, y = success_score)) + geom_point() + geom_smooth(alpha = 0.3) + geom_hline(yintercept = 3)



x = firstRoundWR$weight
y = firstRoundWR$success_score

wr_weight_lm = lm(success_score ~ weight, firstRoundWR)
summary(wr_weight_lm)

firstRoundWR = firstRoundWR %>%
  mutate(predicted_success = -20.47 + .109 * weight, residual = case_when(
    predicted_success <= 0 ~ 0,
    predicted_success > 0 ~ success_score - predicted_success))

ggplot(firstRoundWR, aes(x = weight, y = residual)) + geom_point()

sub200lbFRWR = firstRoundWR %>%
  filter(weight <= 200) %>%
  mutate(suc = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))

above200lbFRWR = firstRoundWR %>%
  filter(weight > 200) %>%
  mutate(suc = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))

firstRoundWR %>%
  filter(success_score >= 3) %>%
  print(n = Inf)
```

```{r}
ggplot(wr, aes(x = height, y = success_score)) + geom_point() + geom_smooth() + geom_hline(yintercept = 3)
ggplot(firstRoundWR, aes(x = height, y = success_score)) + geom_point() + geom_smooth() + geom_hline(yintercept = 3)
```

```{r}
nrow(sub200lbFRWR)
nrow(above200lbFRWR)
```

> Out of 587 wide receivers, 35 have been considered successful (6%)

> Weight is a very big factor in wide receiver success, especially first round wide receivers

> Here is a graph of weight vs success score among all wide receivers drafted in the first round

```{r}
ggplot(firstRoundWR, aes(x = weight, y = success_score)) + geom_point(alpha = 0.3) + geom_smooth() + geom_hline(yintercept = 3)
```

> Here are two box plots showing successful vs non-successful first round wide receivers for those above and below 200 pounds

```{r}
ggplot(sub200lbFRWR, aes(x = suc)) + geom_bar()
ggplot(above200lbFRWR, aes(x = suc)) + geom_bar()
```

```{r}
#test for significance
n = 45
p = .166
mean = n*p
sd = sqrt(mean (1 - p))

pval = (1 - pnorm(12, mean, sd))
pval

```

> While it seems obvious that in football, the bigger players will be more successful. However, there have been 35 wide receivers drafted in the first round that are sub-200 pounds, and 49 that are above 200 pounds, a noticeable difference but clearly teams are comfortable drafting smaller wide receivers in the first round
> Out of the 49 wide receivers drafted in the first round above 200 pounds, 13 have been successful (26.5%)
> While out of the 35 wide receivers drafted in the first round that are sub-200 pounds, only one was successful (2.9%)
> There is extremely strong evidence (pval < 1e-6, one sided test) that a first round receiver that weighs over 200 pounds will be more successful than the average first round wide receiver

> RUNNING BACKS:

```{r}
rb = data %>%
  filter(position == 'RB')

ggplot(rb, aes(x = pick, y = success_score)) + geom_point(alpha = 0.3) + geom_smooth() + geom_hline(yintercept = 3)

top50rbs = rb %>%
  filter(pick <= 50) %>%
  mutate(suc = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))

notTop50 = rb %>%
  filter(pick > 50) %>%
  mutate(suc = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))
```

> Clearly from this graph it shows that draft pick is a massive factor in determining how successful a running back will be in the NFL. However, unlike quarterbacks, if we narrow the sample to only to top50 running backs, we find that other factors contribute as well. In our sample, we have 360 total running backs with 31 being successful (8.6%)

```{r}
ggplot(notTop50, aes(x = suc)) + geom_bar()
ggplot(top50rbs, aes(x = suc)) + geom_bar()
```

> There are 71 running backs who were drafted top 50 and 18 were successful. Of running backs were drafted top 50, 25.3% were successful.

```{r}
top50rbsSpeedFit = top50rbs %>%
  filter(speed <= 4.6, speed >= 4.4) %>%
  mutate(suc = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))
  
speedfit = rb %>%
  filter(speed <= 4.6, speed >= 4.4) %>%
  mutate(suc = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))
```

> However, we can take this a step farther and show that among these top50 running backs, speed plays a huge factor. 

```{r}
ggplot(top50rbs, aes(x = suc)) + geom_bar() + ggtitle("all top 50 rbs")
ggplot(speedfit, aes(x = suc)) + geom_bar() + ggtitle("all rbs 4.4 - 4.6")
ggplot(top50rbsSpeedFit, aes(x = suc)) + geom_bar() + ggtitle("top 50 rbs with 4.4 - 4.6")
```

> 261/360 of all running backs (72.5%) run between a 4.4 and 4.6 forty yard dash. Looking at our top 50 rbs, 52/71 run between a 4.4 and 4.6 forty yard dash (73.2%). So, the proportions are not much different, it seems that running between a 4.4 and 4.6 does not influence where a player gets drafted. But, when you combine the two, there is clear correlation.

```{r}
# probability that a rb BOTH: drafted top 50 AND between 4.4 - 4.6 forty more likely to succeed that those without both
otherRb = rb %>%
  filter((pick > 50 | speed < 4.4 | speed > 4.6) & success_score >=3)

nrow(otherRb)

n = 52
p = .086
mean = n * p
sd = sqrt(mean ( 1 - p))

pval1 = 1 - pnorm(14, mean, sd)
pval1

#top50nonspeedFit %>%
  #filter(success_score >= 3)

#nrow(top50nonspeedFit)

n = 52
p = .086
mean = n * p
sd = sqrt(mean ( 1 - p))

pval2 = 1 - pnorm(14, mean, sd)
pval2

betweenspeed = rb %>%
  filter(speed <= 4.6, speed >= 4.4, success_score >= 3)
nrow(betweenspeed)

n = 52
p = .086
mean = n * p
sd = sqrt(mean ( 1 - p))

pval3 = 1 - pnorm(14, mean, sd)
pval3

nonSpeedFit = rb %>%
  filter(speed < 4.4 | speed > 4.6) %>%
  filter(success_score >= 3)

nrow(nonSpeedFit)

n = 261
p = .086
mean = n * p
sd = sqrt(mean ( 1 - p))

pval4 = 1 - pnorm(24, mean, sd)
pval
```

> Out of our 360 running backs, 308 of them did not meet both of the criteria: drafted top 50 and between 4.4-4.6 forty. Only 5.2% were successful. There is extremely strong evidence (pval 0, one sided test) that a running back who meets both those criteria is more likely to be successful than one who doesn't. Taking it a step further, there is extremely strong evidence (pval 0, one sided test) that a running back who meets both those criteria is more likely to be successful than a running back taken outside the top 50 but running between a 4.4 and 4.6 yard forty. Taking it one more step further, there is still extremely strong evidence (pval < 1e-13, one sided test) that a running back who was drafted top50 is more likely to be successful if they run between a 4.4 and 4.6 forty than a running back who was drafted top50 but doesn't run between a 4.4 and 4.6 forty.

> TIGHT ENDS:

```{r}
te = data %>%
  filter(position == 'TE')

top100te = te %>%
  filter(pick <= 100)
```

> Out of 239 tight ends in our sample, 12 were considered successful (5%)

```{r}
ggplot(top100te, aes(x = pick, y = success_score)) + geom_point(alpha = 0.3) + geom_smooth() + geom_hline(yintercept = 3)
```

> Clearly, draft pick has an influence. 101 tight ends were picked in the top 100 and 11 of them were sucessful (10.7%). Out of tight ends not picked in the top 100 only 1 out of 138 were succesful(0.7%). 

```{r}
top100te = top100te %>%
  mutate(suc = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))

ntop100te = te %>%
  filter(pick > 100) %>%
  mutate(suc = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))

ggplot(ntop100te, aes(x = suc)) + geom_bar() + ggtitle("not top 100")
ggplot(top100te, aes(x = suc)) + geom_bar() + ggtitle("top 100")
```

> Given this disparity, we are going to look exclusively at tight ends drafted in the top 100. Unlike other positions, we can keep going and create a very nice mold for player success using all 3 of height, weight, and speed. 

> The notable criteria is that the tight must be: between 6'5-6'6, 250-265 lbs, and 4.5 - 4.8 forty time.

```{r}
criteria = top100te %>%
  filter(height %in% c("6-5", "6-6"), weight >= 250, weight <= 265, speed >= 4.5, speed <= 4.8)

criteriaALL = te %>%
  filter(pick > 100, height %in% c("6-5", "6-6"), weight >= 250, weight <= 265, speed >= 4.5, speed <= 4.8)

nrow(criteriaALL)

```

> It might seem quite obvious that a player who is big (6'5 or 6'6, 250-265 pounds) AND fast (sub 4.8 at that size) will be more successful and this is a factor as to why they were drafted top 100 in the first place. However, 48 tight ends met that criteria and 27 of them were drafted top 100 (56.3%). This shows that more tight ends that fit this criteria end up going top 100 in the draft than those drafted outside the top 100, but the disparity is not significant. Here the disparity is shown

```{r}
c = te %>%
  filter(height %in% c("6-5", "6-6"), weight >= 250, weight <= 265, speed >= 4.5, speed <= 4.8) %>%
  mutate(top_100 = case_when(
    pick <= 100 ~ 'YES',
    pick > 100 ~ 'NO'
  ))

ggplot(c, aes(x = top_100)) + geom_bar()
```

> Now here is that same disparity, but with the proportion that were successful

```{r}
criteria = criteria %>%
  mutate(suc = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))

criteriaALL = criteriaALL %>%
  mutate(suc = case_when(
    success_score >= 3 ~ 'YES',
    success_score < 3 ~ 'NO'
  ))

ggplot(criteria, aes(x = suc)) + geom_bar() + ggtitle("top 100")
ggplot(criteriaALL, aes(x = suc)) + geom_bar() + ggtitle("top 100")
```

> Here you can see that of the 27 tight ends that met the criteria drafted top 100, 7 were succesful. Of the 21 that were not drafted top 100, 0 were succesful. So, being big and fast is NOT necessarily a factor on its own, but becomes a significant factor among the top 100 draft picks. 

> Now we will test for significance against combinations that don't meet this criteria

> Tight ends that meet the other3 criteria but were not drafted top 100: p val 0

```{r}
df1 = te %>%
  filter(pick <= 100, weight >= 250, weight <= 265, speed >= 4.5, speed <= 4.8, height %in% c("5-11", "6-0", "6-1", "6-2", "6-3", "6-4", "6-7", "6-8"), success_score >= 3)

#1/24 successful

n = 27
p = 1 / 24
mean = n * p
sd = sqrt(mean (1 - p))
pval = 1 - pnorm(6, mean, sd)
pval
```

> Tight ends that meet the other 3 criteria but were not 6'5 or 6'6: p val < 1 e -6

```{r}
df2 = te %>%
  filter(pick <= 100, (weight < 250 | weight > 265), speed >= 4.5, speed <= 4.8, height %in% c("6-5", "6-6"), success_score >= 3)

#1/8 successful

n = 27
p = 1 / 8
mean = n * p
sd = sqrt(mean (1 - p))
pval = 1 - pnorm(6, mean, sd)
pval

```

> Tight ends that meet the other 3 criteria but were not between 250-265 lbs: p val .002

```{r}
df3 = te %>%
  filter(pick <= 100, weight >= 250, weight <= 265, (speed < 4.5 | speed > 4.8), height %in% c("6-5", "6-6"), success_score >= 3)

#1/7 successful

n = 27
p = 1 / 7
mean = n * p
sd = sqrt(mean (1 - p))
pval = 1 - pnorm(6, mean, sd)
pval

```

> Tight ends that meet the other 3 criteria but did not run between a 4.5 and 4.8 40: p val .01

> P values were less than .01 for tight ends meeting 3/4 of those criteria but missing one, for all 4 of the one criteria being missed. This shows that the combination of all 3 criteria is key for predicting tight end success.

> Summary:
> For quarterbacks, there seemed to be little correlation between height/weight/speed and success in the NFL. The only correlation was between draft picks, which was strong. Quarterbacks had a much higher chance of being successful in the NFL if they were drafted in the top 100, and even higher if they were drafted in the first round (top 32).
> For wide receivers, draft pick and weight seem to be the biggest factors in leading to NFL success. First round wide receivers that weigh 200+ pounds have a very strong likelihood of becoming succesful. 
> For running backs, draft pick and speed seem to be the biggest factors in leading to NFL success. Running backs drafted top 50 that run between a 4.4 and 4.6 forty yard dash have a very strong likelihood of becoming successful.
> For tight ends, draft pick, weight, height, and speed are all important factors in leading to NFL success. Tight ends drafted top 100 that are 6'5 or 6'6, between 250-265 pounds, and run between a 4.5 and 4.8 forty have a very strong likelihood in leading to NFL success.